---
title: "PredictionAssignment"
author: "Akila Seneviratne"
date: "February 1, 2016"
output: html_document
---

#Synopsis

##Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.

#Download and Read the Data

This section loads the requiered libraries to R and then downloads the training and test data sets if they are not already in a subdirectory called data. Fianlly the training and test data sets are read into R.

```{r, warning=FALSE}
#load requiered R libraries
library(caret)

#Download data
if(!file.exists("Data")) {
  dir.create("Data")
}
if(!file.exists("./Data/pml-training.csv")) {
  trainUrl<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
  download.file(trainUrl, destfile = "./Data/pml-training.csv")
}
if(!file.exists("./Data/pml-testing.csv")) {
  testUrl<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
  download.file(testUrl, destfile = "./Data/pml-testing.csv")
}

#Read data
training<-read.csv("./Data/pml-training.csv")
validate<-read.csv("./Data/pml-testing.csv")
```

#Data Preprocessing

Many of the features have "NA" entries. These features were first removed from the training data set.

```{r}
training<-training[,(colSums(is.na(training))==0)]
```

Then the features such as X, user_name, raw_timestamp_part_1 etc which have no relationship with the activity of lifting a dumbell was also removed from the training set.

```{r}
#Remove unrelated features from the training set
Index <-  which(names(training) %in% c('X', 'user_name', 'raw_timestamp_part_1', 'raw_timestamp_part_2', 'cvtd_timestamp', 'new_window','num_window'))

training <- training[, -Index]
```

Finally the features which have zero variance was also removed since they will not contribute to the model.

```{r}
trainNzv <- nearZeroVar(training, saveMetrics=TRUE)
training <- training[,trainNzv$nzv==FALSE]
```

# Data Partitioning

Training set was partitioned as training and testing data.

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
train <- training[inTrain,]
test <- training[-inTrain,]
```

#Model Building

Three different models were developed with decision trees, generalized boosted regression and random forests.

```{r, cache=TRUE, warning=FALSE, message=FALSE}
#Model based on decision trees
Model_rpart<-train(classe~.,data=train, method="rpart")
Model_rpart

#Model based on generalized boosted regression
control<-trainControl(number=5)
Model_gbm<-train(classe~.,data=train, method="gbm", trControl=control)
Model_gbm

#Model based on random forests
control<-trainControl(number=5)
Model_rf<-train(classe~.,data=train, method="rf", trControl=control)
Model_rf
```

#Model Testing

The three models were tested using the test data set and by compearing the accuracy of the estimates.

```{r}
#Test the decision tree model
pre_rpart<-predict(Model_rpart,test)
confusionMatrix(pre_rpart, test$classe)

#Test the generalized boosted regression model
pre_gbm<-predict(Model_gbm,test)
confusionMatrix(pre_gbm, test$classe)

#Test the random forests model
pre_rf<-predict(Model_rf,test)
confusionMatrix(pre_rf, test$classe)
```
The decision tree model performed worst of all three models with an accuracy of 0.5. The best model is the random forest model with an accuracy of 0.99 

#Model Validation

Following is the formula to validate the model and to prepare the requiered files to submit as answers for this project
```{r}
validateing<-predict(Model_rf, validate)
validateing
```

When these results were entered in the Course Project Prediction Quiz the answers were 100% match. Therefore the random forest model is the best model for this data set.